---
title: "CLINICAL STUDY OF STROKE"
author: "WAKASALA NOBERT"
date: "2025-02-07"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load the necessary libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(performance)
library(ggeffects)
library(ggcorrplot)
library(equatiomatic)
library(sjPlot)
library(gtsummary)
library(report)
library(vip)
library(car)
library(pROC)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
data <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/Stroke.csv")
glimpse(data)
anyNA(data)
duplicated(data) %>% table()
```

## Data Preparation
```{r warning=FALSE, message=FALSE}
# Convert categorical variables to factors
categorical_vars <- c("hypertension", "heart_disease", "stroke")
data[categorical_vars] <- lapply(data[categorical_vars], as.factor)
data <- data %>% mutate_if(is.character, as.factor)
```

# Convert bmi to numeric
```{r warning=FALSE, message=FALSE}
data$bmi <- as.factor(data$bmi)
data$bmi <- as.numeric(data$bmi)
```

# Create bmi, age and glucose Levels for intuitive study
```{r warning=FALSE, message=FALSE}
data$BMI <- ifelse(data$bmi<18.5, "Underweight",
                       ifelse(data$bmi>=18.5&data$bmi<25, "Normal Weight",
                              ifelse(data$bmi>=25&data$bmi<30, "Overweight",
                                     ifelse(data$bmi>=30&data$bmi<40, "Obese", "Severely Obese"))))
data$Age <- ifelse(data$age <= 30, "Young",
                   ifelse(data$age > 30&data$age<=60, "Middle Age", "Old"))
data$glucose <- ifelse(data$avg_glucose_level >=70&data$avg_glucose_level<=100, "Normal", "Abnormal")
```

## Univariate Analysis
## Summary statistics for numeric variables
```{r warning=FALSE, message=FALSE}
data %>% select(age, avg_glucose_level, bmi) %>% summary()
```

## Majority of the participants were middle aged
## Majority of the participants had an average glucose level of 106.15 which is slightly abnormal.
## Majority of the participants were severely obese which pose them at high risk of developing stroke.


## Frequency distributions for numeric variables
## Theme set
```{r warning=FALSE, message=FALSE}
theme_set(theme_classic()+theme(title = element_text
                           (face = "bold", 
                             colour = "steelblue")))
```

## Age
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(age))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  ggtitle("Distribution of Participants based on Age")

ggplot(data, aes(age))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Participants based on Age")
```

## Average glucose level
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(avg_glucose_level))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  ggtitle("Distribution of Participants based on Average glucose level")

ggplot(data, aes(avg_glucose_level))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Participants based on Average glucose level")
```


## BMI
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(bmi))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  ggtitle("Distribution of Participants based on BMI")
```


## Frequency distributions for categorical variables
## Age groups
```{r warning=FALSE, message=FALSE}
table(data$Age)
ggplot(data, aes(Age))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on Age")
```

## Majority of the participants were middle aged followed by old then young


## BMI Levels
```{r warning=FALSE, message=FALSE}
table(data$BMI)
ggplot(data, aes(BMI))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on BMI")
```

## Majority of the participants were severely obese followed by obese, underweight, overweight and then normal weight.

## Glucose levels
```{r warning=FALSE, message=FALSE}
table(data$glucose)
ggplot(data, aes(glucose))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on glucose")
```

## Participants who had normal glucose level were slightly higher compare to those who had abnormal blood glucose level

## Gender
```{r warning=FALSE, message=FALSE}
table(data$gender)
ggplot(data, aes(gender))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on Gender")
```

## Majority of the participants were female follewed by male then the other genger.

## Hypertention
```{r warning=FALSE, message=FALSE}
table(data$hypertension)
ggplot(data, aes(hypertension, fill = hypertension))+
  geom_bar()+
  ggtitle("Distribution of participants based on Hypertention")+
  scale_fill_discrete(name = "hypertension", labels = c("Negative", "Positive"))
```

## 498 participants had high blood pressure while 4612 did not have.

## Heart Disease
```{r warning=FALSE, message=FALSE}
table(data$heart_disease)
ggplot(data, aes(heart_disease, fill = heart_disease))+
  geom_bar()+
  ggtitle("Distribution of participants based on heart disease")+
  scale_fill_discrete(name = "heart_disease", labels = c("Negative", "Positive"))
```

## Few participants had heart disease compared to those who did not have.

## Ever Married
```{r warning=FALSE, message=FALSE}
table(data$ever_married)
ggplot(data, aes(ever_married))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on Marital status")
```

## Majority of the participants had once in there lifetime been married.

## Work type
```{r warning=FALSE, message=FALSE}
table(data$work_type)
ggplot(data, aes(work_type))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on work type")
```

## Majority of the participants were working privately followed by self employed, government jobs e.t.c.

## Residence type
```{r warning=FALSE, message=FALSE}
table(data$Residence_type)
ggplot(data, aes(Residence_type))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on Residence type")
```

## Majority of the population were living in urban areas compared to those in rural areas.

## Smoking status
```{r warning=FALSE, message=FALSE}
table(data$smoking_status)
ggplot(data, aes(smoking_status))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on smoking status")
```


## Majority of the participants have not in there lifetime smoked.

## Stroke
```{r warning=FALSE, message=FALSE}
table(data$stroke)
ggplot(data, aes(stroke))+
  geom_bar(fill = "purple", color = "red")+
  ggtitle("Distribution of participants based on stroke")
```

## Majority of the participants did not have stroke compared to those who had. This indicates class imbalance which may affect our research study of stroke. 

## Bivariate Analysis
## Correlation

```{r warning=FALSE, message=FALSE}
cr <- data %>% select(age, bmi, avg_glucose_level) %>% cor()
ggcorrplot(cr,title = "Correlation Matrix",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))
```

## There was a weak positive relationship among the numeric variables indicating no autocorrelation and multicollinearity.

## Chi-square
## Is there any relationship between Gender and stroke
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$gender)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is no statistically significant association between gender and stroke indicted by a p-value greater than 0.05.

## Is there any relationship between stroke and hypertension
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$hypertension)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## The test yielded a Chi-square statistic of 81.605 representing how much the observed data deviates from what would be expected if there were no relationship between the variables being studied. The high value indicates a strong association between the two variables. The extremely low p-value indicates a high significant association between the variables. This suggests that there is a substantial connection between hypertension and stroke.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(hypertension, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs Hypertension")
```


## Is there any association between stroke and heart disease
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$heart_disease)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is a strong statistically significant association between stroke and heart disease evidenced by the p-value less than 0.05 and a higher chi-square value. This suggests that heart disease is closely linked to stroke.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(heart_disease, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs Heart Disease")
```

## Is there any association between stroke and smoking status
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$smoking_status)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is a strong statistically significant association between stroke and smoking status evidenced by smaller p-value less than 0.05 and a higher chi-square value.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(smoking_status, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs Smoking status")
```


## Is there any relationship between stroke and marital status
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$ever_married)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is a statistically significant association between stroke and marital status evidenced by a p-valu less than 0.05 and a higher chi-square value. This strongly suggests that marital status is substantially connected to stroke.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(ever_married, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs marital status")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between stroke and working status
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$work_type)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is a statistically significant association batween work type and stroke enidenced by a p-value less than 0.05 and a higher chi-square value. This suggests that work type has a substantial influence on stroke.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(work_type, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs working status")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between stroke and Residence type
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$Residence_type)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is no statistically significant association batween residential type and stroke evidenced by a higher p-value greater than 0.05 and a lower chi-square statistic.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Residence_type, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs Residence type")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between stroke and Age groups
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$Age)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is a statistically significant association between age groups and stroke evidenced by a p-value less than 0.05 and a higher chi-square value. 


```{r warning=FALSE, message=FALSE}
## Visualize your results
ggplot(data, aes(Age, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs Age groups")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between stroke and BMI levels
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$BMI)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```

## There is no statistically significant association between BMI levels and stroke evidenced by a higher p-value greater than 0.05 and a smaller chi-square value.

## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(BMI, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs BMI Levels")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between stroke and Glucose levels
```{r warning=FALSE, message=FALSE}
contingency_table <- table(data$stroke, data$glucose)
contingency_table
chi_square <- chisq.test(contingency_table)
print(chi_square)
```


## There is a strong statistically significant association between glucose levels and stroke evidenced by a smaller p-value less than 0.05 and a higher chi-square statistic.


## Visualize your results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(glucose, fill = stroke))+
  geom_bar(position = "dodge")+
  ggtitle("Stroke vs Glucose Levels")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## T-test
## Is there any relationship between age and stroke
```{r warning=FALSE, message=FALSE}
t.test(age~stroke, data = data)
```

## The p-value associated with age is very small indicating a strong statistical significance. This suggests that there is a statistically significant difference in age between stroke status. Different stroke status are likely to have different age groups.

## Visualize our results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(stroke, age, fill = stroke))+
  geom_boxplot()+
  ggtitle("Age vs Stroke")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between bmi and stroke
```{r warning=FALSE, message=FALSE}
t.test(bmi~stroke, data = data)
```


## The p-value associated with bmi is very small indicating a strong statistical significance. This suggests that there is a statistically significant difference in bmi between stroke status. Different stroke status are likely to have different bmi levels.

## Visualize our results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(stroke, bmi, fill = stroke))+
  geom_boxplot()+
  ggtitle("BMI vs Stroke")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Is there any relationship between average glucose level and stroke
```{r warning=FALSE, message=FALSE}
t.test(avg_glucose_level~stroke, data = data)
```


## The p-value associated with average glucose level is very small indicating a strong statistical significance. This suggests that there is a statistically significant difference in average glucose level between stroke status. Different stroke status are likely to have different average glucose levels.

## Visualize our results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(stroke, avg_glucose_level, fill = stroke))+
  geom_boxplot()+
  ggtitle("Avg glucose level vs Stroke")+
  scale_fill_discrete(name = "stroke", labels = c("Negative", "Positive"))
```


## Multivariate Analysis
## Data Preparation
```{r warning=FALSE, message=FALSE}
data$Age <- NULL
data$BMI <- NULL
data$glucose <- NULL
data$id <- NULL
```


## Implement the logistic regression model
```{r warning=FALSE, message=FALSE}
model <- glm(stroke~., data = data, family = binomial)
summary(model)
```


## Optimize the model
```{r warning=FALSE, message=FALSE}
modela <- glm(stroke~age+hypertension+work_type+avg_glucose_level+bmi, data = data, family = binomial)
```


## Checkmodela## Check model assumtions
## Goodness of fit
```{r warning=FALSE, message=FALSE}
check_predictions(modela) %>% plot()
```


## The models predicted values includes the observed data points.

## Multicollinearity
```{r warning=FALSE, message=FALSE}
check_collinearity(modela)
```


## Homodesdasticity
```{r warning=FALSE, message=FALSE}
check_residuals(modela) %>% plot()
```

## The residuals are uniformly distributed.

## Outliers
```{r warning=FALSE, message=FALSE}
check_outliers(modela)
check_outliers(modela) %>% plot()
```


## Check the models predicted probabilities
```{r warning=FALSE, message=FALSE}
ggeffect(model)
```


## Visualize the models predicted probabilities
```{r warning=FALSE, message=FALSE}
ggeffect(modela) %>% plot() %>% sjPlot::plot_grid()
```


## Check for odds ratios
```{r warning=FALSE, message=FALSE}
fancy_table <- tbl_regression(
  modela, exponentiate = TRUE,
  add_pairwise_contrasts = TRUE,
  contrasts_adjust = "bonferroni",
  pairwise_reverse = TRUE,
  pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% 
  add_significance_stars(hide_p = F, hide_se = T,
                         hide_ci = F) %>% 
  bold_p()
fancy_table
```

## For every one unit change in age, the chances of developing stroke increases by 1.08. This difference is statistically significant.
## People with hypertension have 1.37 higher odds of developing stroke compred to those without hypertension.
## People who have government jobs have 0.19 lower odds of developing stroke compared to children.
## People who work privately have 1.18 higher odds of developing stroke compred to government workers.
## People who work privately have 8733 higher odds of developing stroke compred to those not working.
## The average change in odds of developing stroke for one unit change in average glucose level leads to 1.00 change in chances of developing stroke. The difference is statistically significant.
## For every one unit change in bmi, the odds of developing stroke increases by 1.00. The difference is statistically significant.
## Check feature importance
```{r warning=FALSE, message=FALSE}
car::Anova(modela)
vip(modela)
```

## Age and BMI appeared to be the top predictors of stroke while work type the least predictor of stroke.

## How well is the model in making predictions
```{r warning=FALSE, message=FALSE}
roc(stroke~fitted.values(modela), data = data,
    plot = TRUE, legacy.axes = TRUE, 
    print.auc = TRUE, ci = TRUE)
```

## The model had an AUC of 0.853 making it an excellent tool in classification of stroke.